---
title: "Introduction to hvtiPropensityScores"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Introduction to hvtiPropensityScores}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
library(hvtiPropensityScores)
```

## Overview

`hvtiPropensityScores` provides propensity score methods for cardiac surgery
comparative-effectiveness research.  It ports the balancing logic from our SAS
programs into a tidy R API compatible with
[hvtiPlotR](https://ehrlinger.github.io/hvtiPlotR/).

Two balancing strategies are supported:

- **Nearest-neighbour matching** (`ps_match`): 1:1 greedy matching without
  replacement, with optional caliper.
- **Inverse-probability-of-treatment weighting** (`ps_weight`): IPTW for the
  ATE, ATT, or ATC estimand, with optional stabilisation and weight
  winsorisation.

Both functions share a common `ps_data` S3 base class and return an object
whose `$data` slot carries the original data with the match indicator or weight
column appended, and whose `$tables` slot carries standardised mean difference
(SMD) diagnostics.

---

## Sample data

`sample_ps_data()` generates a synthetic cardiac-surgery dataset that mimics a
typical SAS export: two treatment arms (SAVR / TF-TAVR), several continuous and
binary covariates, a fitted propensity score (`prob_t`), and a `match` indicator
initialised to `0`.

```{r}
#| label: sample-data
dta <- sample_ps_data(n = 500, seed = 42)
head(dta)
```

```{r}
#| label: sample-data-summary
table(dta$tavr)
summary(dta$prob_t)
```

The two treatment groups differ on all covariates — younger patients with higher
ejection fraction were more likely to receive TF-TAVR — so the raw SMDs are
non-trivial and the propensity scores have good overlap.

```{r}
#| label: ps-hist
hist(
  dta$prob_t,
  breaks = 30,
  main   = "Propensity score distribution",
  xlab   = "Estimated probability of TF-TAVR",
  col    = "steelblue",
  border = "white"
)
```

---

## Propensity score matching

### Build the matched object

`ps_match()` performs greedy 1:1 nearest-neighbour matching without
replacement.

```{r}
#| label: match-build
m <- ps_match(dta)
print(m)
```

### Diagnostics

`summary()` prints the available diagnostic tables.

```{r}
#| label: match-summary
summary(m)
```

The `$tables` slot is a named list you can access directly:

```{r}
#| label: match-tables
m$tables$smd_before
m$tables$smd_after
```

### Extract the matched dataset

`$data` is the *full* dataset with a `match` column set to `1` for matched
pairs.  Filter to `match == 1` to obtain the matched subset.

```{r}
#| label: match-data
matched <- m$data[m$data$match == 1L, ]
nrow(matched)
table(matched$tavr)
```

### Caliper matching

A caliper restricts matches to pairs whose propensity scores differ by no more
than the supplied threshold, reducing the number of matched pairs in exchange
for tighter balance.

```{r}
#| label: match-caliper
m_cal <- ps_match(dta, caliper = 0.05)
print(m_cal)
```

---

## Inverse-probability-of-treatment weighting

### ATE weights (default)

`ps_weight()` computes IPTW weights and appends them to `$data`.  The default
estimand is the **average treatment effect** (ATE).

```{r}
#| label: weight-ate
w_ate <- ps_weight(dta, estimand = "ATE")
print(w_ate)
```

```{r}
#| label: weight-ate-summary
summary(w_ate)
```

The `iptw` column holds the stabilised weights.

```{r}
#| label: weight-ate-head
head(w_ate$data[, c("id", "tavr", "prob_t", "iptw")])
summary(w_ate$data$iptw)
```

### ATT and ATC weights

For the **average treatment effect on the treated** (ATT), treated patients
receive weight `1` and controls receive `ps / (1 - ps)`.

```{r}
#| label: weight-att
w_att <- ps_weight(dta, estimand = "ATT")
summary(w_att$data$iptw)
```

For the **average treatment effect on the controls** (ATC), the roles are
reversed.

```{r}
#| label: weight-atc
w_atc <- ps_weight(dta, estimand = "ATC")
summary(w_atc$data$iptw)
```

### Weight trimming (winsorisation)

Extreme weights can destabilise estimates.  Supplying `trim` winsorises weights
to the (`trim`, `1 - trim`) quantile range.

```{r}
#| label: weight-trim
w_trim <- ps_weight(dta, estimand = "ATE", trim = 0.05)
summary(w_trim$data$iptw)
```

---

## Downstream use with hvtiPlotR

The `$data` and `$tables` slots are designed for direct consumption by
`hvtiPlotR::hv_mirror_hist()` (weighted mode) and `hvtiPlotR::hv_balance()`.

### Mirror histogram (matched)

```{r}
#| label: mirror-hist-match
#| eval: false
library(hvtiPlotR)

mh <- hv_mirror_hist(
  data          = m$data,
  x             = prob_t,
  treatment     = tavr,
  match_col     = match,
  group_labels  = c("SAVR", "TF-TAVR")
)
plot(mh) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::labs(x = "Propensity Score", y = "Number of Patients") +
  hv_theme("manuscript")
```

### Mirror histogram (weighted)

```{r}
#| label: mirror-hist-weight
#| eval: false
mh_wt <- hv_mirror_hist(
  data          = w_ate$data,
  x             = prob_t,
  treatment     = tavr,
  weight_col    = iptw,
  group_labels  = c("SAVR", "TF-TAVR")
)
plot(mh_wt) +
  ggplot2::scale_y_continuous(labels = abs) +
  ggplot2::labs(x = "Propensity Score", y = "Weighted Count") +
  hv_theme("manuscript")
```

---

## Session info

```{r}
#| label: session-info
sessionInfo()
```
